---
title: "smchd1 long read MommeD1 only analysis"
author: "Xueyi Dong"
date: "27/09/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
load("countsPrimary.RData")
# counts_run2 <- c1
# load("../Charity/counts/counts.RData")
# counts_run1 <- c1
```

```{r}
library(edgeR)
library(limma)
# source("~/analysis/voom_group/voomByGroup.R")
# x <- DGEList(counts = cbind(counts_run1$counts, counts_run2$counts))
x <- DGEList(counts = fc$counts)
colnames(x)<- c(paste0("BC0", 7:9), paste0("BC", 10:18))
x$samples$group <- factor(c("wt", "fresia", "fresia", "smchd1", "smchd1", "wt", "wt", "fresia", "wt", "wt", "smchd1", "fresia"))
x$samples$run <- factor(rep(c(1, 2), c(6,6)))
x$samples$sample <- c("Fr31", "Fr46", "Fr42", "MD143", "MD142", "MD135", "Fr31", "Fr15", "Fr17", "MD114", "MD140", "Fr46")
x$samples

# remove Fresia samples and low count samples
x <- x[, -c(2, 3, 8, 12)]

filter <- filterByExpr(x)
summary(filter)
x <- x[filter, ]
x <- calcNormFactors(x)
lcpm <- cpm(x, log=TRUE)


x$samples
summary(x$samples$lib.size)

library(RColorBrewer)
col <- brewer.pal(3, "Set1")

pdf("plots/mds.pdf", height = 4, width = 4)
plotMDS(lcpm, col=col[x$samples$group], pch = c(16, 17)[x$samples$run], cex = 1.3)
legend("topright", 
       c("Smchd1-null run 1", "Smchd1-null run 2", "wt run 1", "wt run 2"), 
       pch = c(16, 17, 16, 17), 
       col = col[c(1, 1, 2, 2)], 
       text.col = col[c(1, 1, 2, 2)],
       pt.cex = 1.3)
dev.off()


# plotMDS(removeBatchEffect(lcpm, batch=x$samples$run), col=col[x$samples$group], main="group")
```

### Annotation

```{r annotation}
library(Mus.musculus)
geneid <- rownames(x)
geneid <- substr(geneid, 1, 18)
genes <- select(Mus.musculus, keys=geneid, columns=c("SYMBOL", "TXCHROM","ENSEMBL"), 
                keytype="ENTREZID")
genes <- genes[!duplicated(genes$ENTREZID),]
head(genes)
x$genes <- genes
```

### Design and contrast

```{r}
design <- model.matrix(~0+x$samples$group+x$samples$run)
colnames(design) <- sub("x\\$samples\\$group", "", colnames(design))
colnames(design) <- sub("x\\$samples\\$", "", colnames(design))
cont <- makeContrasts(smchd1-wt, levels=colnames(design))
```

```{r}
# v <- voomByGroup(x, design=design, plot="all")
pdf("plots/voom.pdf", height = 4, width = 6)
v <- voomWithQualityWeights(x, design=design, plot=TRUE, col=col[x$samples$group])
dev.off()
```

### Fit

```{r}
fit <- lmFit(v, design=design)
fit <- contrasts.fit(fit, contrasts=cont)
efit <- eBayes(fit)
dt <- decideTests(efit, p.value = 0.3)
summary(dt)
```

### Top table

```{r toptable}
# for (i in 1:3){
  # cat(colnames(efit)[i], "\n")
  print(topTable(efit, coef=1))
  # cat("\n")
# }
```

### MD plot

```{r MDplot}
pdf("plots/md.pdf")
  plotMD(efit, status=dt, column=1, values=c(1,-1), hl.col=c("red", "green"))
dev.off()
```

### heatmap

```{r}
library(pheatmap)
annotation_col <- data.frame(group = x$samples$group)
rownames(annotation_col) <- colnames(x)
pdf("plots/heatmap.pdf", width = 4)
pheatmap(lcpm[dt!=0,], scale = "row", 
         annotation_col = annotation_col,
         show_rownames = FALSE)
dev.off()
```


## geneset test from Chen et al. 2015
```{r}
up <- read.csv("../Charity/Chen_et_al_2015_PNAS/Upregulated_in_Smchd1_null.csv", skip=1)
down <- read.csv("../Charity/Chen_et_al_2015_PNAS/Downregulated_in_Smchd1_null.csv", skip=1)

index.up <- x$genes$ENTREZID %in% up$EntrezID
index.down <- x$genes$ENTREZID %in% down$EntrezID


plotMD(efit, status=as.numeric(index.up), column=1, main=paste0("Up genes in\n", colnames(efit)[1]))
plotMD(efit, status=as.numeric(index.down), column=1, main=paste0("Down genes in\n", colnames(efit)[1]))
```

### roast
```{r}
index.all <- index.up | index.down
gene.weights <- rep(0, nrow(x))
m <- match(up$EntrezID, x$genes$ENTREZID)
gene.weights[m[!is.na(m)]] <- up$logFC[!is.na(m)]
m <- match(down$EntrezID, x$genes$ENTREZID)
gene.weights[m[!is.na(m)]] <- down$logFC[!is.na(m)]

# for (i in 1:3) {
  cat("Test for Kelan's genes in", colnames(cont)[1], "\n")
  print(roast(v, index.all, gene.weights = gene.weights[index.all], design=design, contrast=cont[,1]))
  cat("\n")
# }

# pdf("barcode.pdf", height=12, width=10)
# png("barcode.png", height = 800, width = 640)
# par(mfrow=c(3,1))
# for(i in 1:3){
  barcodeplot(efit$t[, 1], index=index.all,
              gene.weights = gene.weights[index.all],
              main=colnames(cont)[1])
# }
# dev.off()
```

### Mark Kelan's genes in top table

```{r}
tt <- topTable(efit, coef=1, n=Inf)
tt$status <- "ns"
tt$status[tt$logFC > 0 & tt$adj.P.Val<0.1] <- "up"
tt$status[tt$logFC < 0 & tt$adj.P.Val<0.1] <- "down"
tt$status.chen <- "ns"
tt$status.chen[match(up$EntrezID, tt$ENTREZID)] <- "up"
tt$status.chen[match(down$EntrezID, tt$ENTREZID)] <- "down"
tt$status.all <- paste(tt$status, tt$status.chen, sep="_")
tt$status.all <- as.factor(tt$status.all)
table(tt$status.all)

# write.table(topTable(efit, coef=1, n=Inf), file="results/topTableShortMD.txt", sep='\t')
```

## Gene set test from Chen et al., use less genes

Use FDR cutoff of 0.001, there will be 218 up and 54 down genes.

```{r}
up.strict <- up[up$adj.P.Val < 0.0001, ]
down.strict <- down[down$adj.P.Val < 0.0001, ]
dim(up.strict)
dim(down.strict)
index.up.strict <- x$genes$ENTREZID %in% up.strict$EntrezID
index.down.strict <- x$genes$ENTREZID %in% down.strict$EntrezID
index.strict <- index.up.strict | index.down.strict
# calculate gene weight
gene.weights <- rep(0, nrow(x))
m <- match(up.strict$EntrezID, x$genes$ENTREZID)
gene.weights[m[!is.na(m)]] <- up.strict$logFC[!is.na(m)]
m <- match(down.strict$EntrezID, x$genes$ENTREZID)
gene.weights[m[!is.na(m)]] <- down.strict$logFC[!is.na(m)]
# roast
cat("Test for Kelan's genes in", colnames(cont)[1], "\n")
print(roast(v, index.strict, gene.weights = gene.weights[index.strict], design=design, contrast=cont[,1]))
cat("\n")
```

Up p-value is 0.019. Nice!

Barcode plot

```{r}
pdf("plots/barcodeStrict.pdf", height = 4)
barcodeplot(efit$t[, 1], index=index.strict,
            gene.weights = gene.weights[index.strict],
            main=colnames(cont)[1])
dev.off()
```


### Glimma MD plot

```{r eval=FALSE}
lcpm <- cpm(x, log=TRUE)

levels(tt$status.all)
cols <- c("red", "blue", "lightskyblue", "mediumpurple", "midnightblue", "gray50", "red4", "lightcoral")
library(Glimma)
glMDPlot(efit,  status=tt$status.all[match(rownames(dt), rownames(tt))],
         cols=cols,
         main="Smchd1 vs WT", side.main="SYMBOL",
         counts=lcpm,
         groups = x$samples$group,
         launch = FALSE)
```


## Gene set test using Maayan's short read data

### Roast

```{r}
shortTopTable <- read.table("../../short/resultsDE/topTableShortMD.txt", sep="\t")
index <- x$genes$ENTREZID %in% shortTopTable$ENTREZID[shortTopTable$adj.P.Val<0.1]
m <- match(shortTopTable$ENTREZID[shortTopTable$adj.P.Val<0.1], x$genes$ENTREZID)
gene.weights <- shortTopTable$logFC[na.omit(m)]

cat("Test for Maayan's genes in", colnames(cont)[1], "\n")
print(roast(v, index, gene.weights = gene.weights, design=design, contrast=cont[,1]))
cat("\n")
```

```{r}
  barcodeplot(efit$t[, 1], index=index,
              gene.weights = gene.weights,
              main="Maayan's short read significant genes in long read")
```

## Mark short read results in top table

```{r}
shortDownGenes <- shortTopTable[shortTopTable[m, "adj.P.Val"]<0.1 & shortTopTable[m, "logFC"]<0, "ENTREZID"]
shortUpGenes <- shortTopTable[shortTopTable[m, "adj.P.Val"]<0.1 & shortTopTable[m, "logFC"]>0, "ENTREZID"]
tt$status.short <- "ns"
tt$status.short[tt$ENTREZID %in% shortUpGenes] <- "up"
tt$status.short[tt$ENTREZID %in% shortDownGenes] <- "down"
tt$status.allWithShort <- paste(tt$status.all, tt$status.short, sep="_")
tt$status.LongAndShort <- paste(tt$status, tt$status.short, sep="_")
table(tt$status.LongAndShort)
```

### Compare long read and Maayan's short read log FC 

```{r}
merged <- merge(tt, shortTopTable, by="ENTREZID")
par(mfrow=c(1,2))
smoothScatter(merged$logFC.x, merged$logFC.y, 
              main = paste("logFC, cor=", round(cor(merged$logFC.x, merged$logFC.y), 3), sep=""))
abline(coef = c(0, 1))
smoothScatter(merged$AveExpr.x, merged$AveExpr.y, 
              main=paste("Average expression, cor=", round(cor(merged$AveExpr.x, merged$AveExpr.y), 3), sep=""))
abline(coef = c(0, 1))
```

### Compare long read and old short read log FC

```{r}
tt.chen <- read.table("~/analysis/smchd1/old/TopTable.txt", sep="\t")
m <- match(tt$ENTREZID, tt.chen$EntrezID)
cor(tt$logFC, -tt.chen$logFC[m], use = "complete.obs")

m.up <- match(up.strict$EntrezID, tt$ENTREZID)
m.down <- match(down.strict$EntrezID, tt$ENTREZID)

smoothScatter(tt$logFC, -tt.chen$logFC[m],
              xlab = "long read log FC",
              ylab = "short read log FC",
              )
abline(coef = c(0, 1))
abline(v=0)
abline(h=0)
# highlight significant genes
points(tt$logFC[m.up], -tt.chen$logFC[m][m.up], col="red", pch = 20)
points(tt$logFC[m.down], -tt.chen$logFC[m][m.down], col="green", pch = 20)
```

The correlation is only 0.17. That's low.

Let's compare t.

```{r}
cor(tt$t, -tt.chen$t[m], use = "complete.obs")
smoothScatter(tt$t, -tt.chen$t[m],
              xlab = "long read t-statistics",
              ylab = "short read t-statistics",
              )
abline(coef = c(0, 1))
abline(v=0)
abline(h=0)
# highlight significant genes
points(tt$t[m.up], -tt.chen$t[m][m.up], col="red", pch = 20)
points(tt$t[m.down], -tt.chen$t[m][m.down], col="green", pch = 20)
```

Oh, even lower

## Save top table

```{r}
write.table(topTable(efit, coef=1, n=Inf), file = "results/TopTableMD.txt", sep="\t")
```




